Analyzing Performance Bottleneck in Graph Neural Network Training: An Experimental View

### 2020-07-16 Manuscript/sec4-experiments.tex
update section4

\paragraph{the neccessity of removing abnormal epochs}
The absolute training time of four GNN models is shown in \figurename~\ref{fig:exp_absolute_training_time}.
In figure, we can conclude that the time ranking is GaAN $\gg$ GAT $>$ GGNN $>$ GCN. In all datasets, the number of edges is greater than that of nodes, so the correlation of training time between GAT and GGNN is
GAT $>$ GGNN. And the remaining relationship is consistent with complexity analysis. In addition, we find that there are some outliers in \figurename~\ref{fig:exp_absolute_training_time} because of profiling overhead
and GC pause of the Python interpreter. So in all the experiments below, we will remove abnormal epochs and use the average of remaining epochs to analysis.

@@

\paragraph{the effects of hyper-parameters on training time}

According to \tablename~\ref{tab:gnn_overview_edge}, \tablename~\ref{tab:gnn_overview_vertex}, the computational complexity of edge and vetex of each GNN is linearly related to the 
hyper-parameters of each model(such as $d_{dim}$, $K$, etc at al). 
In order to prove this linear relationship, we measured the change of the training time of each GNN model with the hyperparameters.

@@

@@

For GCN and GGNN, the computational complexity is affected by the hidden vector dimension $d_{dim}$. $d_{dim}$ affects both the output hidden vector dimension of Layer0 and
the input hidden vector dimension of Layer1($d_{dim} = d^0_{out} = d^1_{in}$).
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gcn} and \figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_ggnn}
display that the training time of GCN and GGNN is affected by $d_{dim}$. As $d_{dim}$ increases, the training time increases linearly.

For GAT,  which uses a multi-head mechanism, its computational complexity is affected by the input hidden vector dimension $d_{in}$ and the hidden vector dimension $d_{head}$ of each head, and the number of heads $K$.
The output hidden vector dimension of each layer is $d_{out}=K d_{head}$.
Because of GAT model structure $d^1_{in}=d^0_{out}$, adjusting $d_{head}$ and $K$ is equivalent to adjusting Layer1’s $d^1_{in}$.
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gat} shows that the GAT training time increases linearly with $d_{head}$ and $K$, two.

For GaAN, which also has a multi-head mechanism, its computational complexity is affected by $d_{in}$, $d_v$, $d_a$ and the number of heads $K$.
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gat} demonstrates that GaAN training time is linearly affected by hyperparameters.
\\
The above experiments prove the complexity analysis results given in \tablename~\ref{tab:gnn_overview_edge}, \tablename~\ref{tab:gnn_overview_vertex}. 
The training time of each GNN model increases linearly with the increase of hyperparameters.
When the hidden vector dimension $d_{in}$ is too low, the calculations involving hidden vectors account for a very low proportion of the total calculation time,
resulting in an insignificant change in the total training time.
When the hidden vector dimension is large enough, the total training time increases linearly with $d_{in}$.

\paragraph{the effects of hyper-parameters on memory usage}

\figurename~\ref{fig:exp_hyperparameter_memory_usage} shows how each GNN model's memory usage varies with the algorithm's hyperparameters.
With the increase of hyperparameters, GNN's memory usage also increases linearly.

@@

We can answer the question Q1 \ref{sec:experimental_env}: the complexity analysis in \tablename~\ref{tab:gnn_overview_edge}, \tablename~\ref{tab:gnn_overview_vertex} is valid. 
\textbf{GNN training time and memory usage are linearly related to hyperparameters}.
This allows algorithm engineers to use larger hyperparameters to increase the complexity of GNN without worrying about the time-consuming training and the explosive growth of memory usage.

@@

In thi subsection, we aims to find what stage is the most time-consuming stage in GNN training by decomposing the time-consuming training.

\paragraph{vertex/edge calculation time-consuming proportion analysis}

For vertex calculation and edge calculation, \figurename~\ref{fig:exp_vertex_edge_cal_proportion} shows the ratio of the vertex/edge calculation time of different GNN layers in each GNN
to the total training time (including forward, backward and evaluation stages). 
The GCN algorithm spends most of time on edge calculation in most datasets except \textit{cph} dataset on account of
the dimension of the input feature vector of this dataset is very high, resulting in the high vertex calculation time-consuming of Layer0.
Because of its high vertex calculation complexity, GGNN makes its vertex calculation time-consuming\
ratio significantly higher than other algorithms, but in most datasets, edge calculation still occupies the majority of total time. 
Only in \textit{pub} and cam dataset, the edge calculation cost and the vertex calculation cost are close for GGNN 
because the average degree of the two datasets is low (only 4.5 and 2.8).
For the GAT and GaAN algorithms, because of their high edge calculation complexity, their edge calculation cost time is absolutely dominant. 
In summary, \textbf{edge calculation is the main time-consuming factor of GNN training}, especially in the case of more complicated edge calculation.

@@

The edge calculation phase can be further decomposed into four steps of collect, message, aggregate and update,
as shown in \figurename~\ref{fig:steps_in_edge_calculation}. The figure shows the edge calculation process of GNN layer $l$.
edge index is a matrix that holds the edge sets of the graph with shape $M*2$, where $M$ is the number of edges,
and the two columns of the matrix respectively store the source vertex and the target vertex of each edge.
The edge index remains unchanged throughout the calculation process.
The collect step is used to prepare the data structure required for edge calculation and will use the vertex hidden vector of the GNN layer $\boldsymbol{h}_i^l (1 \leq i \leq N)$
as input, which is copied to the two layers of each edge according to the edge index, forming the input parameter tensor of the edge calculation function $\phi$
(including $\boldsymbol{h}_i^l$,$\boldsymbol{h}_j^ l$ and $\boldsymbol{e}_{j, i}^l$).
There is no calculation in this step, only data access is involved. 
The message step calls the function given by the user $\phi$ to complete the edge calculation process, and get message vector of each edge $\boldsymbol{m}_{j, i}^l (\boldsymbol{e}_{j, i}^l \in E(G))$.
The aggregate step is based on the target of each edge, the message vector with the same target vertex is aggregated by the aggregation operator $\Sigma$,
and then each vertex aggregation vector $\boldsymbol{a}_i^l (1 \leq i \leq N)$ is obtained. 
The last update step is optional, and it can perform additional correction processing on the aggregated vector (for example, adding bias to GCN).
The aggregated vector $a_i^l$ after the update process will be input into the vertex calculation function $\gamma$ as an input parameter.

@@

>>>>

\paragraph{The neccessity of removing abnormal epochs}
The absolute training time of four GNN models is shown in \figurename~\ref{fig:exp_absolute_training_time}.
In figure, we can conclude that the time ranking is GaAN $\gg$ GAT $>$ GGNN $>$ GCN. In all datasets, the number of edges is greater than that of nodes, so the correlation of training time between GAT and GGNN is
GAT $>$ GGNN. And the remaining relationship is consistent with complexity analysis. In addition, we find that there are some outliers in \figurename~\ref{fig:exp_absolute_training_time} because of profiling overhead
and GC pause of the Python interpreter. So in all the experiments below, we will remove abnormal epochs and use the average of remaining epochs to analysis.

@@

@@

\paragraph{The effects of hyper-parameters on training time}
According to \tablename~\ref{tab:gnn_overview_edge}, \tablename~\ref{tab:gnn_overview_vertex}, the computational complexity of edge and vetex of each GNN is linearly related to the 
hyper-parameters of each model(such as $d_{dim}$, $K$, etc at al). In order to prove this linear relationship, we measured the change of the training time of each GNN model with the hyperparameters.
For GCN and GGNN, the computational complexity is affected by the hidden vector dimension $d_{dim}$. $d_{dim}$ affects both the output hidden vector dimension of Layer0 and
the input hidden vector dimension of Layer1($d_{dim} = d^0_{out} = d^1_{in}$).
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gcn} and \figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_ggnn}
display that the training time of GCN and GGNN is affected by $d_{dim}$. As $d_{dim}$ increases, the training time increases linearly.
For GAT,  which uses a multi-head mechanism, its computational complexity is affected by the input hidden vector dimension $d_{in}$ and the hidden vector dimension $d_{head}$ of each head, and the number of heads $K$.
The output hidden vector dimension of each layer is $d_{out}=K d_{head}$.
Because of GAT model structure $d^1_{in}=d^0_{out}$, adjusting $d_{head}$ and $K$ is equivalent to adjusting Layer1’s $d^1_{in}$.
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gat} shows that the GAT training time increases linearly with $d_{head}$ and $K$, two.
For GaAN, which also has a multi-head mechanism, its computational complexity is affected by $d_{in}$, $d_v$, $d_a$ and the number of heads $K$.
\figurename~\ref{fig:exp_hyperparameter_on_vertex_edge_phase_time_gat} demonstrates that GaAN training time is linearly affected by hyperparameters.
These experiments prove the complexity analysis results given in \tablename~\ref{tab:gnn_overview_edge}, \tablename~\ref{tab:gnn_overview_vertex}. 
The training time of each GNN model increases linearly with the increase of hyperparameters.
When the hidden vector dimension $d_{in}$ is too low, the calculations involving hidden vectors account for a very low proportion of the total calculation time,
resulting in an insignificant change in the total training time.
When the hidden vector dimension is large enough, the total training time increases linearly with $d_{in}$.

@@

\paragraph{The effects of hyper-parameters on memory usage}
\figurename~\ref{fig:exp_hyperparameter_memory_usage} shows how each GNN model's memory usage varies with the algorithm's hyperparameters.
With the increase of hyperparameters, GNN's memory usage also increases linearly.

@@

We can answer the question Q1 in \ref{sec:experimental_env}: the complexity analysis in \tablename~\ref{tab:gnn_overview_edge}, \tablename~\ref{tab:gnn_overview_vertex} is valid. 
\textbf{GNN training time and memory usage are linearly related to hyperparameters}.
This allows algorithm engineers to use larger hyperparameters to increase the complexity of GNN without worrying about the time-consuming training and the explosive growth of memory usage.

@@

In this subsection, we aims to find what stage is the most time-consuming stage in GNN training by decomposing the time-consuming training.

\paragraph{Vertex/edge calculation time-consuming proportion analysis}

\figurename~\ref{fig:exp_vertex_edge_cal_proportion} shows the ratio of the vertex/edge calculation time of different GNN layers
to the total training time (including forward, backward and evaluation). GCN spends most of time on edge calculation in most datasets. A special case is  \textit{cph} dataset because of
the dimension of the input feature vector of this dataset is very high, resulting in the high vertex calculation of Layer0.
High vertex calculation makes GGNN's vertex calculation time-consuming ratio significantly higher than other algorithms.
But in most datasets, for GGNN, edge calculation still occupies the majority of total time. 
In \textit{pub} and cam dataset, the edge calculation cost and the vertex calculation cost are close for GGNN 
because the average degree of the two datasets is low (only 4.5 and 2.8).
For the GAT and GaAN algorithms, due to their high edge calculation complexity, their edge calculation cost time is absolutely dominant. 
In summary, \textbf{edge calculation is the main time-consuming factor of GNN training}, especially in the case of more complicated edge calculation.

@@

@@

The edge calculation phase can be further decomposed into four steps of collect, message, aggregate and update,
as shown in \figurename~\ref{fig:steps_in_edge_calculation}. The figure shows the edge calculation process of GNN layer $l$, 
edge index is a matrix that holds the edge sets of the graph with shape $M*2$, where $M$ is the number of edges,
and two columns of the matrix respectively store the source vertex and the target vertex of each edge.
Edge index remains unchanged throughout the calculation process.
The collect step is used to prepare data structure required for edge calculation and will use the vertex hidden vector of the GNN layer $\boldsymbol{h}_i^l (1 \leq i \leq N)$
as input, which is copied to the two layers of each edge according to the edge index, forming the input parameter tensor of the edge calculation function $\phi$
(including $\boldsymbol{h}_i^l$,$\boldsymbol{h}_j^ l$ and $\boldsymbol{e}_{j, i}^l$).
There is no calculation in this step, only data access is involved. 
The message step calls the function given by the user $\phi$ to complete the edge calculation process, and get message vector of each edge $\boldsymbol{m}_{j, i}^l (\boldsymbol{e}_{j, i}^l \in E(G))$.
The aggregate step is based on the target of each edge, the message vector with the same target vertex is aggregated by the aggregation operator $\Sigma$,
and then each vertex aggregation vector $\boldsymbol{a}_i^l (1 \leq i \leq N)$ is obtained. 
The last update step is optional, and it can perform additional correction processing on the aggregated vector (for example, adding bias to GCN).
The aggregated vector $a_i^l$ after the update process will be input into the vertex calculation function $\gamma$ as an input parameter.
